CrystalPlan: an Experiment Planning Tool for Crystallography

### 2010-09-22 docs/CrystalPlan_paper/paper.tex
Refs #182: version of the paper I will send around for comments.

\documentclass[draft]{iucr}              % DO NOT DELETE THIS LINE

@@

\usepackage{graphicx}
\usepackage{gensymb}
\usepackage[version=3]{mhchem} 
%\usepackage[pdftex]{graphicx}
%\DeclareGraphicsExtensions{.pdf,.png,.jpg}

@@

Crystal plan is designed with these capabilities:

\begin{itemize}
  \item Load an instrument geometry.
  \item Calculate the orientation of a crystal at different goniometer settings.
  \item Calculate the coverage of these orientations in reciprocal space.
  \item Take crystal symmetries into account.
  \item Optimize the experiment plan.
\end{itemize}

@@

The reciprocal space to be measured is divided into a cubic 3D matrix, the size
of which is determined by the user, who specifies the minimum d-spacing
($d_{min}$, in $\AA$) that he or she is interested in. The resolution of the 3D
matrix is also specified (in $\AA^{-1}$), which determines the number of points to be
modeled. Available machine memory limits the resolution that can be achieved in
practice.

@@

When entering the lattice parameters, the crystal's point group symmetry can
also be entered. This information is then used to calculate the increased
measurement redundancy. A list of $n$ $3\times3$ multiplication matrices is
generated based on the point group; any $hkl$ vector can be multiplied by
these matrices to find the $n$ equivalent $hkl$ reflections; for example, for
orthorombic (mmm) symmetry, $n=8$ and 8 multiplication matrices are generated, which multiply
$hkl$ to give $\pm h$, $\pm k$, $\pm l$.

@@

As part of the TOPAZ beamline commissioning process, a Natrolite
(\ce{Na2Al2Si3O10} $\cdot$ \ce{2H2O}) single crystal was measured on a fixed
$45\degree\ \chi$ goniometer at room temperature and ambient atmosphere. An initial short run was processed using
ISAW to determine the UB matrix and cell parameters of the crystal.

@@

Peaks were integrated using ISAW, and the resulting .integrate file was loaded
into CrystalPlan for comparison. The predicted and actual measurement
positions on each detector face were compared for the 554 integrated peaks
found. Very good agreement was found; the
RMS difference between the predicted and actual measurement wavelength was only
0.016 $\AA$, whereas the RMS of the detector position error was
7.6 mm. Inspecting the offsets showed that the main contributor to this error was a
slight misalignment of the goniometer, which CrystalPlan can help diagnose.

@@

\begin{figure}
\caption{Volume coverage in reciprocal space view: a) without crystal symmetry;
b) with crystal symmetry}
%\includegraphics{fig1.ps}
\label{fig:volume_view}
\end{figure}

\begin{figure}
\caption{Coverage of single-crystal reflections in reciprocal space: a) without
crystal symmetry; b) with crystal symmetry}
%\includegraphics{fig1.ps}
\label{fig:refl_view}
\end{figure}

>>>>

\documentclass[final]{iucr}              % DO NOT DELETE THIS LINE

@@

\usepackage{graphicx}
\usepackage{gensymb} %General symbols
\usepackage[version=3]{mhchem} %For chemical formulae
\usepackage{subfigure}

%\usepackage[pdftex]{graphicx}
%\DeclareGraphicsExtensions{.pdf,.png,.jpg}

%quick angstrom define
\newcommand{\ang}{$\mathrm{\AA} $}

@@

CrystalPlan is designed with these capabilities:

\begin{itemize}
  \item Load an instrument geometry.
  \item Calculate the orientation of a crystal at different goniometer settings.
  \item Calculate the coverage of these orientations in reciprocal space,
  in volume and for single peaks.
  \item Take crystal symmetries into account.
  \item Optimize the experiment plan automatically.
\end{itemize}

@@

The reciprocal space to be measured is divided into a cubic 3D matrix, the size
of which is determined by the user, who specifies the minimum d-spacing
($d_{min}$, in \ang) that he or she is interested in. The resolution of the
3D matrix is also specified (in \ang$^{-1}$), which determines the number of
points to be modeled. Available machine memory limits the resolution that can be achieved in
practice.

@@

When entering the lattice parameters, the crystal's point group symmetry can
also be entered. This information is then used to calculate the increased
measurement redundancy. A list of $n$ $3\times3$ multiplication matrices is
generated based on the point group; any $hkl$ vector can be multiplied by
these matrices to find the $n$ equivalent $hkl$ reflections; for example, for
orthorombic (mmm) symmetry, $n=8$; therefore 8 multiplication matrices are
generated that multiply $hkl$ to give $\pm h$, $\pm k$, $\pm l$.

@@

As part of the TOPAZ beamline commissioning process, a Natrolite
(\ce{Na2Al2Si3O10} $\cdot$ \ce{2H2O}) single crystal was measured on a fixed
$45\degree\ \chi$ goniometer at room temperature and ambient atmosphere. An
initial short run (wavelength range from $\sim0.5$ \ang\ to $3.5$ \ang)
was processed using ISAW to determine the UB matrix and cell parameters of the crystal.

@@

Peaks were integrated using ISAW, and the resulting .integrate file was loaded
into CrystalPlan for comparison. The predicted and actual measurement
positions on each detector face were compared for the 554 integrated peaks
found. Very good agreement was found; the
RMS difference between the predicted and actual measurement wavelength was only
0.016 \ang, whereas the RMS of the detector position error was
7.6 mm. Inspecting the offsets showed that the main contributor to this error was a
slight misalignment of the goniometer, which CrystalPlan can help diagnose.

@@

\begin{figure}
\caption{Sample screenshot showing the 3D view of the predicted coverage of
reciprocal space volume for 3 sample orientations and 14 detectors in TOPAZ, from 0.9 \ang\ to 4.0\ang:
a) without crystal symmetry, 17\% coverage; b) with crystal symmetry (mmm 8-fold
symmetry), coverage increases to 65\%. }
\includegraphics{volume_nosym.eps}
a)
\includegraphics{volume_sym.eps}
b)
\label{fig:volume_view}
\end{figure}

\begin{figure}
\caption{Sample screenshot showing the 3D view of single-crystal reflection
measurements. Each sphere represents an $hkl$ peak, color-coded to indicate the
number of measurements (dark blue: 0, cyan: 1, green: 2, etc.). 
a) without crystal symmetry, 2375 of 9481 reflections are measured, or 25\%.
b) with crystal symmetry, 1276 of 1427 primary reflections are measured, or
89\%, and redundancy is also increased. }
\includegraphics{refl_nosym.eps}
a)
\includegraphics{refl_sym.eps}
b)
\label{fig:refl_view}
\end{figure}
